Guarded Recursion in Agda via Sized Types

### 2019-01-26 code-oneclock-paper/latex/Introduction.tex
Fixes in introduction

Dependent type theory is a very expressive functional programming
language that underlies the deductive system of proof assistants such
as Coq, Agda and Lean.  It is a total language,
meaning that every program definable inside type theory is necessarily
terminating. This is an important requirement that ensures the
consistency of the type system.

@@

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
\cite{Hughes96} is a type annotated with the number of unfolding that
elements of this type can undergo. Sized types have been implemented
in Agda \cite{A-sized,AVW-normalization} and can be used in
combination with coinductive records for the specification of
coinductive types \cite{Copatterns}.  Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \F{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \F{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus with guarded recursion.
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Sized types and guarded recursion are different solution to the same
problem, but a thorough study of the relation between these two
disciplines is still missing. In this paper we take a first step in
this direction by showing how guarded recursion can be simulated in
Agda using sized types.

Utilizing techniques for representing ``type theory in type theory''
\cite{Chapman09,AltenkirchK16}, we present an Agda formalization of
the syntax of a simple type theory with guarded recursion. This object
language, that we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aformentioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston et al. \cite{CloustonBGB15} have studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant'' .  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. In this respect \GTT\ is
closer in spirit to Atkey and McBride's language. The design of \GTT\
also has the benefit of allowing a more pleasant introduction rule for
\IC{□} than Clouston et al.'s.

Afterwards we develop a categorical semantics
for \GTT\ using sized types. More precisely, we define a presheaf
model where the type formers of simply typed lambda calculus are
evaluated using the standard Kripke semantics. Typically the semantics
of type theories for guarded recursion is given in the topos of
trees or variations thereof \cite{BMSS-synthetic,Mogelberg14,MannaaM18}.
Here, in order to make clear the connections between guarded recursion and
sized types, we take the preorder of sizes as the indexing
category of our presheaves. This means that types and contexts of \GTT\ are
interpreted as antitone sized types. 
The well-order relation on sizes is fundamental for constructing a
terminating definition of the semantic fixpoint combinator.

Abel and Vezzosi \cite{AbelV14} have also formalized a simple type theory
extended with the later modality in Agda. They focus on operational
properties of the calculus and give a certified proof of strong
normalization. In our work we consider a richer type system and we
formalize its denotational semantics in Agda. Nevertheless, we do not
investigate the dynamic behaviour of our calculus, which we leave to
future work.

>>>>

Dependent type theory is an expressive functional programming
language that underlies the deductive system of proof assistants such
as Coq, Agda and Lean.  It is a total language,
meaning that every program definable inside type theory is necessarily
terminating. This is an important requirement that ensures the
consistency of the type system.

@@

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
\cite{Hughes96} is a type annotated with the number of unfolding that
elements of this type can undergo. Sized types have been implemented
in Agda \cite{A-sized,AVW-normalization} and can be used in
combination with coinductive records to specify
coinductive types \cite{Copatterns}.  Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \F{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \F{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus with guarded recursion.
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Sized types and guarded recursion are different solutions to the same
problem, but a thorough study of the relation between these two
disciplines is still missing. In this paper we take a first step in
this direction by showing how guarded recursion can be simulated in
Agda using sized types.

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory with guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aformentioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant'' .  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. In this respect \GTT\ is
closer in spirit to Atkey and McBride's language. The design of \GTT\
also has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

Afterwards we develop a categorical semantics
for \GTT\ using sized types. More precisely, we define a presheaf
model where the type formers of simply typed lambda calculus are
evaluated using the standard Kripke semantics. Typically the semantics
of type theories for guarded recursion is given in the topos of
trees or variations thereof \cite{BMSS-synthetic,MannaaM18,Mogelberg14}.
Here, to clarify the connection between guarded recursion and
sized types, we take the preorder of sizes as the indexing
category of our presheaves. This means that types and contexts of \GTT\ are
interpreted as antitone sized types. 
The well-order relation on sizes is fundamental for constructing a
terminating definition of the semantic fixpoint combinator.

Abel and Vezzosi \cite{AbelV14} also formalized a simple type theory
extended with the later modality in Agda. They focus on operational
properties of the calculus and give a certified proof of strong
normalization. In our work we consider a richer type system and we
formalize its denotational semantics in Agda. Nevertheless, we do not
investigate the dynamic behaviour of our calculus, which we leave to
future work.
